---
title: "project"
author: "Eric Friesth, Aashwin Lamsal, Ethan Wanlass"
date: "11/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gun Violence incidences recorded in the US from 2013 to 2018

### Data Cleaning

```{r}
#Initial cleaning was done using MS Excel, we dropped columns that we weren't planning to use (Address, house district, senate district, part2-5_age) and removed rows with empty cells.
data = read.csv('gun-violence-202.csv')
head(data)
str(data)


#Changing data type for date from factor to date

data$date <- as.Date(data$date, format = "%m/%d/%Y")
#head(data)
#Date format is in YYYY-MM-DD

#Changing the data type for age from factor to int
data$part1_age <- as.integer(data$part1_age)
#head(data)
#str(data)

#Next, we drop rows with NA values using na.omit.

data_clean <- na.omit(data)

# Finally, we noticed in our two factor variables, state and city_or_county, had levels "" (empty strings) that had to be removed as they were unwanted.

# We did this using the following code to drop levels.

data_clean$state[data_clean$state == ""] = NA
data_clean$state = droplevels(data_clean$state)

data_clean$city_or_county[data_clean$city_or_county == ""] = NA
data_clean$city_or_county = droplevels(data_clean$city_or_county)

# Extracted the year and month from the date to be used for more analysis

data_clean$year = substring(data_clean$date, 1, 4)

data_clean$month = substring(data_clean$date, 6, 7)

#changing the data types from characters to numeric

data_clean$year <- as.numeric(data_clean$year)
data_clean$month <- as.numeric(data_clean$month)

# Final view of cleaned data

head(data_clean)
str(data_clean)

```

## Data Exploration

```{r, echo=FALSE}
library(tidyverse)
population = read.csv('population.csv')
head(population)

unused_names = c("United States","Northeast Region","South Region","West Region", "Puerto Rico")
clean_population = population  %>% select(NAME,POPESTIMATE2018) %>% 
  filter(NAME %in% data_clean$state)%>% droplevels(clean_population$Name)

colnames(clean_population)[1] = "state"
str(clean_population)

```

The population data was gathered from the U.S. Census Bureau, taken from https://www.census.gov/data/tables/time-series/demo/popest/2010s-state-total.html. In this data set we deleted the unnecessary rows (regions and Puerto Rico) as well as Columns not in use. Additionally, the column for the states was called named, so it was renamed to state to match that of the kaggle set. 



## Data Exploration continued

### Number of Incidents per State

One of our interests in this data set was finding the states with the most incidents and seeing in those states where those incidents were occurring. Since, we're using a sample of a sample, the results will not be the direct number of those incidents, but it is proportional. To find this, a table of the count of each state was found as well as a graph ranking the states from the most incidents to least incidents.

```{r, echo=FALSE}
sort(table(data_clean$state), decreasing = TRUE)

incidents_per_state = data_clean %>% 
  count(state) %>%
  group_by(state)

incidents_per_state
levels(incidents_per_state$state)

ordered_incidents_per_state = incidents_per_state
ordered_incidents_per_state$state= reorder(incidents_per_state$state, -incidents_per_state$n)

ggplot(ordered_incidents_per_state, aes(state, n)) + geom_bar(stat = "identity", colour = "black", fill = "rosy brown") + theme(axis.text.x = element_text(angle=90, hjust=1)) + labs(title = "Number of Incidents per State", x = "Number of Incidents", y = "State")
```

Looking at these two figures, we see that Illinois, California and Florida have the top three number of incidents. Basing code from Randy Lai from https://stackoverflow.com/questions/23206834/how-to-plot-a-us-state-in-r , we created the plots of those incidents for Illinois, California and Florida. 

```{r, echo=FALSE}
#Shooting locations for Illinois

Illinois = filter(data_clean, state == "Illinois")
Illinois_map = map_data('state', region = 'Illinois')

ggplot() + geom_polygon(data = Illinois_map, aes(x=long, y=lat,group=group),colour="black", fill="white" )+  geom_point(data = Illinois,aes(x=longitude,y=latitude),alpha = .2,colour="red")+ coord_fixed()

#Shooting locations for California

California = filter(data_clean, state == "California")
California_map = map_data('state', region = 'California')

ggplot() + geom_polygon(data = California_map, aes(x=long, y=lat,group=group),colour="black", fill="white" )+  geom_point(data = California,aes(x=longitude,y=latitude),alpha = .2,colour="red") + coord_fixed()

#Shooting locations for Florida

Florida = filter(data_clean, state == "Florida")
Florida_map = map_data('state', region = 'Florida')

ggplot() + geom_polygon(data = Florida_map, aes(x=long, y=lat,group=group),colour="black", fill="white" )+  geom_point(data = Florida,aes(x=longitude,y=latitude),alpha = .2,colour="red") + coord_fixed()
```

The incidents for each of these states are mostly occurring around the major cities in that State. For Illinois, most of the incidents are in the upper right of the state where chicago. For California, the major hotspots are around their coastal cities: San Francisco, Los Angeles and San Diego. For Florida, the major hotspots are around Miami, Orlando, Tampa and Jacksonville. Thus, it appears that the metropolitan areas have the most incidents. 

Another factor to consider is the influence of population into these incidents. Illinois, California and Florida all have high state population in the U.S., so we wanted to see the relation of population to the number of incidents.  

```{r, echo=FALSE}
ggplot(per_capita, aes(POPESTIMATE2018,n)) + geom_point() + labs(title = "2018 Predicted Population to Number of Incidents", x = "Population", y = "Number of Incidents")
```

Looking at the plot, one can see there is a moderate to strong positive relation between the two. Thus, we'd like to compare the number of incidents per capita, using the predicted population for 2018, to see if there's any change from before. Similarly, the per capita numbers are not going to be exact, but should be proportional. 

```{r, echo=FALSE}
per_capita = inner_join(incidents_per_state,clean_population, by = "state")
per_capita = per_capita %>% mutate(inc_per_capita = n/POPESTIMATE2018)  
str(per_capita)

per_capita$state = reorder(per_capita$state,-per_capita$inc_per_capita)
ggplot(per_capita,aes(state,inc_per_capita)) + geom_bar(stat = "identity", colour = "black", fill = "plum")  + theme(axis.text.x = element_text(angle=90, hjust=1)) + labs(title = "Incidents per Capita per State", y= "Incidents per Capita", x = "State")
```

Illinois is still is in the top three of the incidents per capita, but California and Florida has dropped significantly. Thus, it appears that those states have more incidents due to the population size. Now we'd like to see where the incidents occurred for the top two states incidents per capita, Delaware and Louisiana, based off of the same code from Randy Lai. 

```{r, echo=FALSE}

Delaware = filter(data_clean, state == "Delaware")
Delaware_map = map_data('state', region = 'Delaware')


ggplot() + geom_polygon(data = Delaware_map, aes(x=long, y=lat,group=group),colour="black", fill="white" )+  geom_point(data = Delaware,aes(x=longitude,y=latitude),alpha = .2,colour="red") + coord_fixed()

#Shooting locations for Louisiana

Louisiana = filter(data_clean, state == "Louisiana")
Louisiana_map = map_data('state', region = 'Louisiana')

ggplot() + geom_polygon(data = Louisiana_map, aes(x=long, y=lat,group=group),colour="black", fill="white" )+  geom_point(data = Louisiana,aes(x=longitude,y=latitude),alpha = .2,colour="red") + coord_fixed() 
```

These maps seem simlar to our results earlier, with the major hot spots being situated by the major cities. However, there are significantly less incidences than those from earlier.  



## Data Visualization

### Heatmap
#### Our heatmap is saved in our github repo as 'data_clean.png'
```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(sf)
library(mapview)
library(RColorBrewer)
```

```{r, echo=FALSE, include=FALSE}

#Creation of our continental US Map
map_bounds = c(left = -125, bottom = 25, right = -65, top = 50)

data_clean.map = get_stamenmap(map_bounds, zoom = 7, maptype = "toner-lite")

data_clean.map = ggmap(data_clean.map, extent = "device", legend = "none")

data_clean.map = data_clean.map + stat_density2d(data=data_clean, aes(x = data_clean$longitude, y = data_clean$latitude, fill = ..level.., alpha = ..level..), geom="polygon")

data_clean.map = data_clean.map + scale_fill_gradientn(colours = rev(brewer.pal(7, "Spectral")))


data_clean.map = data_clean.map + theme_bw()

# the code snippet below adds red points over each gun violence incident to the previous map
#data_clean.map <- data_clean.map + geom_point(data=data_clean,  aes(x=data_clean$longitude, y=data_clean$latitude), fill="red", shape=23, alpha=0.8)

ggsave(filename = "D:\\Documents\\ds202_project\\data_clean.png") #change this to something else for your own local directories

```

### Pairplot/Correlation Matrix
```{r}

# Scatter Pair plot:

pairs(data_clean[ ,7:18], pch = 19, lower.panel = NULL)

```

### Actor age vs. The sum of number killed between 2013 and 2018
```{r, echo=FALSE}
# Histogram: part1_age x sum(n_killed)

ggplot(data_clean, aes(x = part1_age, y = sum(n_killed))) + geom_histogram(stat = "identity", fill = "blue") + labs(x="Ages of Actors", y = "Total Number Killed", title = "Total Killed by Actors of Each Particular Age (2013-2018)")

```

### Distribution of Gun Violence Incidents over Time (2013-2018)
```{r, echo=FALSE}

# Histogram: year

ggplot(data_clean, aes(x = year)) + geom_histogram(stat = "count",  fill = "#42a14b") + labs(x = "Year", y = "Frequency", title = "Distribution of Incidents by Year")
```

### Distribution of Gun violence incidents over Time (months)
```{r, echo=FALSE}
# Histogram: month

ggplot(data_clean, aes(x = month)) + geom_histogram(stat = "count", fill = "#42a14b") + labs(x = "Month", y = "Frequency", title = "Distribution of Incidents by Month") + ylim(0, 1200)
```

### Distributions of actors who were arrested vs. actors who were killed, Nationally compared to Chicago, IL

```{r, echo = FALSE}
#National
ggplot(data_clean, aes(x = month, y = n_part_arrested)) + geom_bar(stat= "identity", fill = "#FB8B24") + labs(title = "Number of Actors Arrested by Month in the U.S.", x = "Month", y = "Number of Actors Arrested") + ylim(0, 650)
ggplot(data_clean, aes(x = month, y = n_part_killed)) + geom_bar(stat= "identity", fill = "#FB8B24") + labs(title = "Number of Actors Killed by Month in the U.S.", x = "Month", y = "Number of Actors Killed") + ylim(0, 650)

# Chicago
data_clean_chi = subset(data_clean, data_clean$city_or_county == "Chicago")

ggplot(data_clean_chi, aes(x = month, y = n_part_arrested)) + geom_bar(stat="identity", fill = "#05668D") + ylim(0, 25) + labs(title = "Number of Actors Arrested by Month in Chicago", x = "Month", y = "Number of Actors Arrested")

ggplot(data_clean_chi, aes(x = month, y = n_part_killed)) + geom_bar(stat="identity", fill = "#05668D") + ylim(0, 25) + labs(title = "Number of Actors Killed by Month in Chicago", x = "Month", y = "Number of Actors Killed")

```

## Predictive Models
### Teen actors in relation to the number killed/injured
```{r}
#Predictive Models below
#First, we look at Linear Models

#Looking for relationships between the number of teen actors and any of the other variables 
fit0 <- lm(n_teen_part~n_killed+n_injured,data=data_clean)
summary(fit0)
#Results show an Adjusted R-squared of 0.01936, an F-statistic of 107 with a p-value < 2.2e-16
#This can be interpreted that the estimate of the mean number of teen actors is 0.097279 when n_killed and n_injured are both 0.
#The 0.01 estimate for n_killed means that the mean value of teenage actors would be estimated to increase by 0.01 if n_killed increased by 1 unit, holding all other covariates constant.
#The 0.085432 estimate for n_injured means that the mean value of teenage actors would be estimated to increase by 0.085432 if n_injured increased by 1 unit, holding all other covariates constant.

#However, these results show that the number of people killed or injured are not good predictors for estimating the number of teenaged actors in gun violence, as it would be too difficult to predict something as unpredictable as gun violence.

```

### Number of people killed in relation to the different age groups of actors.

```{r}
#Looking for relationships between n_killed and the different age groups of participants 
fit1 = lm(n_killed~n_adult_part+n_teen_part+n_child_part, data=data_clean)
summary(fit1)

# UPDATE THIS COMMENT LATER:   Results to be interpreted in a similar manner as ABOVE

```

### The number of actors arrested, in relation to the age of the actor, disregarding children.

```{r}
fit2 = lm(n_part_arrested~n_adult_part+n_teen_part, data=data_clean)
summary(fit2)
#Results show an Adjusted R-squared of 0.1643, an F-statistic of 1056 with a p-value < 2.2e-16
#This can be interpreted that the estimate of the mean number of teen actors is -0.012580 when n_adult_part and n_teen_part are both 0. This means that 0.012580 of the average person is being unincarcerated when the number of teen and adult actors are both equal to 0. This doesn't make sense since you'd imagine the estimate to be 0 when the covariates are 0.
#The 0.330548 estimate for n_adult_part means that the mean value of actors arreseted would be estimated to increase by 0.330548 if n_adult_part increased by 1 unit, holding all other covariates constant.
#The 0.393224 estimate for n_teen_part means that the mean value of arrested actors would be estimated to increase by 0.393224 if n_teen_part increased by 1 unit, holding all other covariates constant.
#The correlation isn't strong enough to draw meaningful conclusions from this model, as the R squared is only 16%


```

### The number of actors killed vs the number of victims killed

```{r}
fit3 = lm(n_part_killed~n_adult_part+n_teen_part+n_killed, data=data_clean)
summary(fit3)

#Results show an Adjusted R-squared of 0.9963, an F-statistic of 9.562e+05 with a p-value < 2.2e-16
#This can be interpreted that the estimate of the mean number of actors who've died is 0.0011 when n_adult_part, n_teen_part, and n_killed are both 0 (though it's not 0, it's very close). This means that 0.0011 of the average actor is dead when the number of teen, adult actors, and number killed are all equal to 0. 
#The 0.0002468 estimate for n_adult_part means that the mean value of actors killed would be estimated to increase by 0.0002468 if n_adult_part increased by 1 unit, holding all other covariates constant.
#The -0.0009959 estimate for n_teen_part means that the mean value of dead actors would be estimated to increase by -0.0009959 if n_teen_part increased by 1 unit, holding all other covariates constant (this doesn't make much sense as the opposite of dying is coming back to life, which goes to show that these weren't good predictors).
# The 0.9925804 estimate for n_killed means taht the mean value of dead actors would increase by almost 1 when n_killed increases by 1 unit, holding all covariates constant.
#The correlation is strong enough, with an adjusted R-squared of 0.9963, but this dataset never clarified whether or not the actor gets added to this count whenever he/she dies. It seems like that might be the case due to the R-squared value being almost 1.


```



######################################################################
            
              Changes to be made formatting-wise
              
######################################################################

# Put each plot in a separate R block DONE

# Title each plot with ### "TItle" DONE

# Each plot block needs echo=FALSE DONE

# Give each section a header

# Move results and conclusion over from slides to below regressions DONE

  EXPOUND UPON THE FOLLOWING:
  This given dataset was not geared towards predictive modeling, but the variables themselves are interesting and easy to draw insights from.
Through the geographic information, we can tell that there are some areas with much more violence than others.
Demographic information showed that actors were, typically, in their late teens to mid twenties, and male.
Teen actors are more likely to be arrested than killed by police.
We did not have any data on the ethnicities of the people who were killed, that could lead to further insights.
Summer months are the most active for gun violence.
High rates of incidents occurring in January as well.
Illinois is the state with the most incidents of gun violence, with Chicago being the most violent city.
This insight was made after the sampling of the data, beforehand, Illinois was 3rd, trailing Texas and California
Most incidents of gun violence ARE NOT MASS SHOOTINGS!
A typical incident doesn’t even have deaths involved.
Gun violence is on the rise.
2018’s subtotal shows around 13,000 incidents had occurred, and we only had information for up until March.












